---
title: 'Exercises 2: Flights at ABIA'
author: "Brooks Beckelman, Zack Bilderback, Davis Townsend"
output: html_document
---

```{r echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

```{r}
library(mosaic)
library(foreach)
library(lubridate)
library(ggplot2)
library(plyr)

rm(list=ls())
flights = read.csv("abia.csv")
attach(flights)
```

Our goal in exploring this data set is to provide analysis that allows one to make choices that optimize his or her air travel experience. This means discovering the best and worst times of year to fly, the best and worst times of day to fly, and the best and worst airlines to fly with.

## Time of Year

```{r}
flights_dep = flights[flights$Origin == 'AUS',] # Flights departing from ABIA
flights_arr = flights[flights$Dest == 'AUS',] # Flights arriving at ABIA
flights_DepDelay = flights_dep[flights_dep$DepDelay>0,] # Dataframe of flights with departure delays
flights_ArrDelay = flights_arr[flights_arr$ArrDelay>0,] # dataframe of flights with arrival delays

# Find fraction of flight departures that were delayed in each month
months=c(1:12)
num_of_DepDelays = rep(0,length(months))
num_of_ArrDelays = rep(0,length(months))
num_of_flights = rep(0,length(months))
num_of_deps = rep(0,length(months))
num_of_arrs = rep(0, length(months))
for (month in months){
  DepDelay_month = flights_DepDelay[flights_DepDelay$Month==month,]
  ArrDelay_month = flights_ArrDelay[flights_ArrDelay$Month==month,]
  num_of_DepDelays[month] = nrow(DepDelay_month)
  num_of_ArrDelays[month] = nrow(ArrDelay_month)
  deps_month = flights_dep[flights_dep$Month==month,]
  arrs_month = flights_arr[flights_arr$Month==month,]
  num_of_deps[month] = nrow(deps_month)
  num_of_arrs[month] = nrow(arrs_month)
}

fraction_DepDelay = num_of_DepDelays/num_of_deps
fraction_ArrDelay = num_of_ArrDelays/num_of_arrs

plot(months, num_of_deps, type='b', col='blue', main='Flights by Month', xlab='Month', ylab='Number of Flights', xaxt='n')
points(months, num_of_arrs, col='gold4')
lines(months, num_of_arrs, col='gold4')
axis(side=1, at=months, labels=months)
legend("bottomleft", c("Departures","Arrivals"), lty=c(1,1), col=c('blue','gold4'))

plot(months, num_of_DepDelays, type='b', col='blue', main='Delays by Month', xlab='Month', ylab='Number of Delayed Flights', xaxt='n', ylim=c(1600,3000))
points(months, num_of_ArrDelays, col='gold4')
lines(months, num_of_ArrDelays, col='gold4')
axis(side=1, at=months, labels=months)
legend("bottomleft", c("Departures", "Arrivals"), lty=c(1,1), col=c('blue','gold4'))

plot(months, fraction_DepDelay, type='b', col='blue', main='Fraction of Flights Delayed by Month', xlab='Month', ylab='Fraction of Flights Delayed', xaxt='n', ylim=c(0.4,0.75))
points(months, fraction_ArrDelay, col='gold4')
lines(months, fraction_ArrDelay, col='gold4')
axis(side=1, at=months, labels=months)
legend("topleft", c("Departures", "Arrivals"), lty=c(1,1), col=c('blue', 'gold4'))
```

These plots outline flights on a monthly basis. It seems that the best time to fly to avoid delays is between September and November. December appears to be the month when a flight is most likely to be delayed.

---

## Time of Day

```{r}
# Find fraction of flight departures that were delayed each hour
hours = c(0:23)
timeBeg = 0
num_of_deps_ToD = rep(0, length(hours))
num_of_DepDelays_ToD = rep(0, length(hours))
num_of_arrs_ToD = rep(0, length(hours))
num_of_ArrDelays_ToD = rep(0, length(hours))
for (hr in hours){
  timeInt = timeBeg:(timeBeg+99)
  dep_timeInt = flights[flights$CRSDepTime %in% timeInt,]
  arr_timeInt = flights[flights$CRSArrTime %in% timeInt,]
  num_of_deps_ToD[hr] = nrow(dep_timeInt)
  num_of_arrs_ToD[hr] = nrow(arr_timeInt)
  DepDelay_timeInt = flights_DepDelay[flights_DepDelay$CRSDepTime %in% timeInt,]
  ArrDelay_timeInt = flights_ArrDelay[flights_ArrDelay$CRSArrTime %in% timeInt,]
  num_of_DepDelays_ToD[hr] = nrow(DepDelay_timeInt)
  num_of_ArrDelays_ToD[hr] = nrow(ArrDelay_timeInt)
  timeBeg = timeBeg+100
}

fraction_DepDelay_ToD = num_of_DepDelays_ToD/num_of_deps_ToD
fraction_ArrDelay_ToD = num_of_ArrDelays_ToD/num_of_arrs_ToD

plot(hours, num_of_deps_ToD, type='b', col='blue', main='Flights by Scheduled Departure/Arrival Time', xlab='Hour', ylab='Number of Flights', xaxt='n')
points(hours, num_of_arrs_ToD, col='gold4')
lines(hours, num_of_arrs_ToD, col='gold4')
axis(side=1, at=hours[seq(0,length(hours),2)], labels=hours[seq(0,length(hours),2)])
legend("topright", c("Departures", "Arrivals"), lty=c(1,1), col=c('blue', 'gold4'))

plot(hours, num_of_DepDelays_ToD, type='b', col='blue', main='Delays by Scheduled Departure/Arrival Time', xlab='Hour', ylab='Number of Delays', xaxt='n')
points(hours, num_of_ArrDelays_ToD, col='gold4')
lines(hours, num_of_ArrDelays_ToD, col='gold4')
axis(side=1, at=hours[seq(0,length(hours),2)], labels=hours[seq(0,length(hours),2)])
legend("topright", c("Departures", "Arrivals"), lty=c(1,1), col=c('blue', 'gold4'))

plot(hours, fraction_DepDelay_ToD, type='b', col='blue', main='Fraction of Flights Delayed by Scheduled Departure/Arrival Time', xlab='Hour', ylab='Fraction of Flights Delayed', xaxt='n', ylim=c(0,0.5))
points(hours, fraction_ArrDelay_ToD, col='gold4')
lines(hours, fraction_ArrDelay_ToD, col='gold4')
axis(side=1, at=hours[seq(0,length(hours),2)], labels=hours[seq(0,length(hours),2)])
legend("topleft", c("Departures","Arrivals"), lty=c(1,1), col=c('blue','gold4'))
```

These plots break down flights by scheduled departure and arrival times on an hourly basis. Hour 0 corresponds to all flights departing/arriving between 12 AM and 12:59 AM, hour 1 corresponds to flights between 1 AM and 1:59 AM, etc. It seems that flights scheduled to depart early in the morning and late at night are the least likely to be delayed. However, there are a limited number of flights late at night. 

---

## Airlines
```{r}
#Setup

#dataframe of frequency of total flights by carrier
airplane = read.csv("ABIA.csv", header=TRUE)
attach(airplane)

carriercount = count(airplane$UniqueCarrier)
colnames(carriercount) <- c("UniqueCarrier","TotalFlights")
```

When examining airlines, we will consider two factors: delays and cancellations.

<br>

### Delays

```{r}
#Delays

#first create new dataframe combining total flight count and total delay time
x = data.frame(airplane$UniqueCarrier, airplane$CarrierDelay)
y = aggregate(cbind(airplane$CarrierDelay) ~ UniqueCarrier, data=x, FUN=sum)
colnames(y) <- c("UniqueCarrier","TotDelayTime")

new = data.frame(carriercount$UniqueCarrier, carriercount$TotalFlights, y$TotDelayTime)

#define colr of columns vector for use in graphs
cols <- c('black', 'black', 'black', 'blue','blue', 'blue', 'blue','blue', 
          'blue', 'blue','blue', 'blue', 'gold3', 'gold3', 'gold3', 'gold3')

#now plot total delay time per carrier in a graph
newsorted <- new[order(new$y.TotDelayTime), ]
revdfsorted <- newsorted[rev(rownames(newsorted)),]

barplot(revdfsorted$y.TotDelayTime, names.arg=revdfsorted$carriercount.UniqueCarrier, col=cols, ylim=c(0,70000), 
        ylab = 'Total Delay Time (in minutes)', xlab = 'Airlines', main = 'Total Time Delay of Flights by Airline', las=2)

#now calculate average delay time
new["AvgTimeDelay"] <- new$y.TotDelayTime / new$carriercount.TotalFlights

#sort the airlines by delay time percentage
test <- new[order(new$AvgTimeDelay), ]
revdf <- test[rev(rownames(test)),]

#now plot this in a bar graph
#Highlight airlines with most and least avg of delayed flights
barplot(revdf$AvgTimeDelay, names.arg=revdf$carriercount.UniqueCarrier, col=cols, ylim=c(0,10), 
        ylab = 'Average Time Delay (in minutes)', xlab = 'Airlines', main = 'Average Time Delay of Flights by Airline',
        las=2)
## We see that Mesa Airlines, Inc.(YV) has the highest average delay times, while Envoy Air(MQ) has the least average delay time
```

The preceding plots show the total delay time for each airline in 2008 and the average delay time per flight for each airline. The average was taken over all flights, including those that were not delayed, to give a better idea of how long one can expect to be delayed for a flight from that airline. There are a clear three airlines with longer average delays, Mesa Airlines (YV), ExpressJet (EV), and PSA airlines (OH). There are a clear four airlines with shorter average delays, Envoy Air (MQ), US Airways (US), Southwest Airlines (WN), and Frontier Airlines (F9).

<br>

### Cancellations

```{r}
#Cancellations

#which carrier had most cancellations
x1 = data.frame(airplane$UniqueCarrier, airplane$Cancelled)
y1 = aggregate(cbind(airplane$Cancelled) ~ airplane$UniqueCarrier, data=x1, FUN=sum)
colnames(y1) <- c("UniqueCarrier","TotalCancellations")

#define colr of columns vector for use in graphs
cols <- c('black', 'blue', 'blue', 'blue','blue', 'blue', 'blue','blue', 
          'blue', 'blue','blue', 'blue', 'blue', 'gold3', 'gold3', 'gold3')

#first create new dataframe combining total flight count and total cancellations
new1 = data.frame(carriercount$UniqueCarrier, carriercount$TotalFlights, y1$TotalCancellations)

#now plot total delay time per carrier in a graph
new1sorted <- new1[order(new1$y1.TotalCancellations), ]
revdfsorted1 <- new1sorted[rev(rownames(new1sorted)),]

barplot(revdfsorted1$y1.TotalCancellations, names.arg=revdfsorted1$carriercount.UniqueCarrier, col=cols, ylim=c(0,600), 
        ylab = 'Total Cancellations', xlab = 'Airlines', main = 'Total Cancellations by Airline', las=2)

#now find and plot average cancellations per carrier
new1["AvgCancellation"] <- new1$y1.TotalCancellations / new$carriercount.TotalFlights

#sort by Average Cancellation
test1 <- new1[order(new1$AvgCancellation), ]
revdf1 <- test1[rev(rownames(test1)),]

#now plot this in a bar graph
#Highlight airlines with most and least avg of delayed flights
barplot(revdf1$AvgCancellation, names.arg=revdf1$carriercount.UniqueCarrier, col=cols, ylim=c(0,.1), 
        ylab = 'Average Cancellation Percent', xlab = 'Airlines', main = 'Average Cancellation Rate by Airline', las=2)

# Envoy Air(MQ) had the highest cancellation rate, while Northwest Airlines (NW) had the lowest cancellation rate at 0!
```

These plots illustrate the total cancellations and average cancellation rate for each airline. American Airlines (AA) has almost three times as many cancellations as any other airline, but Envoy Air (MQ) has the highest average cancellation rate. Northwest Airlines (NW), Frontier Airlines (F9), and US Airways (US) have the best cancellation rates. 